\section{Introduction}\label{introduction}

Business process simulation provides a way to analyse process redesigns
before changes are introduced in an organisation. Staffing rules,
routing policies, automation levels, and escalation procedures can be
evaluated by generating hypothetical executions under explicit modelling
assumptions. In process mining, simulation is particularly useful
because historical event logs already record activities, resources,
timestamps, case arrivals, and observed variants. A data-driven
simulator can therefore be calibrated from the same logs used for
process discovery and conformance checking.

Despite this empirical grounding, many BPS approaches remain more
process-centric than resource-centric. They represent the process as a
control-flow structure with timing and resource parameters attached to
it. This representation supports many what-if analyses, but it can
obscure the decentralised character of organisational work. Individual
resources differ in capabilities, workload, availability, handover
behaviour, and local routines. Cases progress because persons, teams, or
software resources accept work, prioritize it, perform it, and transfer
it to other actors.

Recent agent-based approaches respond to this limitation by representing
resources as agents. AgentSimulator, for example, moves data-driven BPS
toward an agent-based view by learning agent behaviour from event logs
\cite{kirchdorfer2024agentsimulator}. In parallel, work on
LLM-empowered agent-based modelling argues that LLMs can provide richer
agent cognition, communication, and decision-making in simulations
\cite{gao2024llmabm}. Together, these lines of work suggest a concrete
research opportunity: a business process can be simulated as a set of
log-grounded resource agents, with LLM-style modules used for local
resource-assignment and handover decisions rather than for free-form
process generation.

This opportunity also introduces a methodological risk. If LLM agents
freely invent process behaviour, the resulting simulation may be
plausible in language while being invalid as process data. For
process-mining applications, each simulated action must remain
compatible with an event-log schema, a feasible activity-resource
assignment, and a measurable quality framework. This study therefore
treats LLM agents as constrained decision modules. They operate within
action spaces derived from event logs, and their generated event logs
remain subject to the same log-distance measures used for conventional
BPS.

The current prototype focuses on the resource-assignment and handover
layer. Activity occurrences are generated from log-derived trace
behaviour, while the compared policies decide which feasible resource
performs each activity and how work moves between resource agents. A
process-mining-compatible simulated log therefore means an event log
with the required case, activity, resource, start-time, and end-time
fields, with activity-resource assignments restricted to feasible
combinations and with outputs that can be evaluated by BPS log-distance
measures.

The main research question is:

\textbf{How can LLM-style local decision modules be embedded into a
resource-centric, log-derived business process simulator without
breaking event-log compatibility?}

This question is examined through two subquestions. First, how can
resource capabilities, timing samples, and handover priors constrain
local resource-assignment decisions? Second, how should such a simulator
be evaluated both as a generated event-log model and as an auditable
agent-behaviour system?

The contribution is a design-science artifact and an initial controlled
evaluation. The artifact, denoted as \(H_{R-LLM}\),
represents resources as agents with learned capabilities, service-time
distributions, local resource priors, handover priors, and memory. It
compares three executable policies: a centralised baseline, a
log-derived agent-profile policy, and a constrained LLM-agent proxy. The
LLM-agent proxy is a deterministic, inspectable stand-in for the
decision interface that a real LLM would occupy. The artifact also
includes an optional API-backed LLM condition behind the same
action-masked interface. The reported quantitative experiments use the
proxy so that the architecture can be evaluated reproducibly before
introducing prompt nondeterminism, latency, and cost.

Compared with AgentSimulator-style behaviour learners, the proposed
artifact makes the local decision interface explicit. Feasible resources
are action-masked, log-derived priors are exposed to the policy,
decisions produce machine-readable resource assignments, and LLM-agent
conditions additionally emit reasoning and handover traces. The study
therefore extends resource-centric BPS by adding a constrained local
decision interface and trace outputs, rather than by replacing
log-derived process simulation.

The evaluation uses two public event logs: AcademicCredentials from the
artifact accompanying Chapela-Campa et al.'s BPS quality framework
\cite{chapela2025quality}, and LoanApp from AgentSimulator
\cite{kirchdorfer2024agentsimulator}. Simulated logs are compared with
held-out logs using lightweight prototype metrics and formal BPS
log-distance measures; what-if intervention experiments additionally
examine scenario response. The results show that the LLM-agent proxy does not
dominate statistical baselines, but remains competitive on selected
log-quality dimensions, improves workforce-distribution distance in
both datasets, and provides additional reasoning and handover
observability. These findings motivate a dual evaluation perspective
combining BPS log quality with LLM-agent behavioural quality.
