\section{Discussion}\label{discussion}

The study began from a simple but risky idea: use LLM-style agents to
simulate business processes. The results show that the idea becomes
scientifically useful only after it is narrowed. A business process
simulator cannot rely on unconstrained language generation. It needs
log-derived feasible action spaces, timing distributions, resource
priors, and validation metrics. Once those constraints are in place, an
LLM-style decision module can be treated as one component inside a
larger BPS architecture rather than as the whole simulator.

The first implication is that resource-centric modeling is a strong
foundation for LLM-agent BPS. The agent-profile baseline performs best
on several formal metrics, which suggests that much of the process
behavior can already be captured by careful log-derived resource
profiles. This is not a problem for the proposed approach; it is a
design lesson. LLM modules should be added where they provide
incremental value, such as ambiguous handovers, policy-sensitive
routing, exception handling, or explanation generation. They should not
replace statistical components that already reproduce observed behavior
well.

The second implication is that LLM-agent simulation should be evaluated
multidimensionally. The LLM-agent proxy looks favorable under some
lightweight metrics, but the formal repeated Chapela-Campa evaluation
shows a more mixed picture. This difference confirms the value of using
a mature BPS quality framework. A single score or a single hand-picked
example would be misleading. The artifact should be judged by how it
changes control-flow, resource, and temporal behavior simultaneously.
At the same time, the absence of uniform outperformance under BPS
log-quality metrics should not be interpreted as evidence that
LLM-agent simulation has no added value. These metrics primarily reward
historical distribution reproduction. They do not fully measure whether
an agent selected only feasible actions, produced a faithful rationale,
handled handovers coherently, adapted to an intervention, or left an
auditable decision trace. This is why the proposed dual framework
separates BPS log quality from LLM-agent behavioral quality.

The third implication concerns interpretability. The LLM-agent proxy
emits reasoning and handover traces that conventional event logs do not
contain. These traces are not a substitute for quantitative validation,
but they can support diagnosis and governance. For example, if an
LLM-agent simulator overuses a historically strong resource, the
reasoning log can reveal whether the policy is repeatedly selecting the
resource because of capability priors, workload estimates, or handover
preferences. This creates a pathway toward auditable agentic process
simulation.

The practical consequence is that future API-backed LLM experiments
should not be judged only by whether they reduce n-gram or cycle-time
distance. They should also report invalid-output rate, fallback rate,
feasible-action compliance, reason-action consistency,
reason-context consistency, handover-message consistency, and scenario
sensitivity. Some of these metrics can be computed deterministically
from the simulator logs, while others require a small human-coded or
LLM-as-judge rubric. In the latter case, judge reliability and bias must
be checked rather than assumed.

Several limitations should be acknowledged. First, the main reported
LLM-agent condition is a deterministic proxy even though the artifact now
contains an optional API-backed LLM implementation. This was intentional
for reproducibility, but it means the reported quantitative comparison
does not yet measure prompt sensitivity, model drift, invalid textual
outputs, or token cost from an archived real-LLM run. Second,
AcademicCredentials remains the primary dataset and LoanApp is used only
as a robustness check. This improves empirical breadth but is still not
enough for broad generalization across process types. Third, the temporal model remains
simple. Although empirical waiting-time and case-arrival sampling
improved fidelity, the simulator does not yet include explicit resource
calendars, multitasking, or data-aware case attributes. Fourth, the
evaluation reports descriptive repeated-run statistics but does not yet
include full significance testing or confidence intervals for all
pairwise comparisons.

Future work should extend the artifact in four directions. The first is
to execute, archive, and evaluate a cost-controlled API-backed
\texttt{llm\_agent\_real} experiment, preferably activating the model
only at ambiguous handover decisions. The second is to add LLM-specific
metrics such as invalid output rate, fallback rate, feasible action-set
size, reason-action consistency, and handover-message consistency. The
third is to evaluate additional logs from the Chapela-Campa and
AgentSimulator artifact families, especially logs with different
resource density and timing characteristics. The fourth is to test intervention scenarios in which
the LLM-agent module may have a clearer advantage, such as changes in
escalation policy or resource specialization that are difficult to
express with static historical priors alone.
