\documentclass[
	a4paper,
	pagesize,
	pdftex,
	12pt,
	ngerman,
	fleqn,
	final,
]{scrartcl}

\usepackage[utf8]{inputenc}
\usepackage[ngerman,english]{babel}
\usepackage[unicode=true]{hyperref}
\usepackage{setspace}
\usepackage{tikz}
\usepackage{tabularx}
\usepackage[draft=false,babel,tracking=true,kerning=true,spacing=true]{microtype}
\microtypesetup{nopatch={footnote}}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{calc}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{float}

\begin{document}
\selectlanguage{english}

\input{Institutsvorlage}

\titel{Resource-Centric LLM-Augmented Multi-Agent Process Simulation for Business Process Mining}
\typ{Study Project Report}
\grad{Master of Science (M. Sc.)}
\autor{Qingzi}
\gebdatum{DD.MM.YYYY}
\gebort{Birthplace}
\gutachter{First Supervisor}{Second Supervisor}
% \mitverteidigung
\makeTitel

\begin{abstract}
\textbf{Abstract.}
Business process simulation (BPS) is used to test process redesigns before they are tried in real organizations. Data-driven BPS can learn control-flow, resource, and timing patterns from event logs, but resources are often treated as centralized pools or parameterized execution units. This study asks whether a more resource-centric simulator can use LLM-style local decisions without losing the reproducibility and measurability required in process mining. I design and evaluate \(H_{R-LLM}\), a log-grounded multi-agent simulation artifact that extracts resource profiles, feasible action spaces, service-time samples, waiting-time samples, and handover priors from event logs. The artifact compares a centralized baseline, a log-derived resource-agent baseline, and a constrained LLM-agent proxy, and it includes an optional API-backed LLM condition behind the same action-masked interface. Experiments on AcademicCredentials and AgentSimulator LoanApp show a consistent trade-off. The statistical baselines remain strongest on several control-flow and timing reproduction measures, while the LLM-agent proxy is competitive and obtains the best workforce Earth Mover's Distance on both datasets. In LoanApp high-load what-if analysis, the proxy also gives the strongest throughput and cycle-time response, but this advantage does not hold under a combined reduced-capacity and high-load stress test. The results therefore do not show that LLM-agent simulation uniformly outperforms traditional BPS. They show instead that LLM-style decision modules can be embedded in a process-mining-compatible simulator, add reasoning and handover traces, and require evaluation as a scenario-dependent combination of log quality and agent-behavior quality.
\end{abstract}
\newpage

\pagenumbering{gobble}
\tableofcontents
\cleardoublepage

\pagenumbering{arabic}
\input{chapters/01_introduction}
\input{chapters/02_background_related_work}
\input{chapters/03_research_design}
\input{chapters/04_artifact_design_implementation}
\input{chapters/05_experimental_setup}
\input{chapters/06_results}
\input{chapters/07_discussion}
\input{chapters/08_code_data_availability}
\input{chapters/09_conclusion}

\newpage
\pagenumbering{roman}
\bibliographystyle{abbrv}
\bibliography{library/citations}

\section*{Appendix}\label{Appendix}
\addcontentsline{toc}{section}{Appendix}
\selbstaendigkeitserklaerung{\today}

\end{document}
